---
title: "Proteomics_analysis"
author: "Ema"
date: "2026-08-10"
output: html_document
---

---
title: "Quantitative Proteomics Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
---

# Overview

## Aim

This analysis demonstrates a reproducible workflow for quantitative proteomics data, starting with protein abundance measurements and progressing through quality control, exploratory data analysis, differential protein abundance analysis, and visualization of temporal responses.

The main objectives are to:

1.  Import and preprocess quantitative proteomics data.
2.  Identify and assess quantitative protein abundance measurements.
3.  Evaluate missing values and sample quality.
4.  Explore sample relationships using correlation analysis and PCA.
5.  Identify differentially abundant proteins using `limma`.
6.  Visualize differential protein abundance using volcano plots.
7.  Compare significant protein sets across multiple time points.

The analysis focuses on a time-course experiment containing control and treatment conditions measured at 30 minutes, 1 hour, 2 hours, and 4 hours.

------------------------------------------------------------------------

# 1. Required R packages

The following packages are used for data manipulation, statistical analysis, quality control, and visualization.

```{r setup, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(dplyr)
library(ggplot2)
library(tidyr)
library(tidyverse)

library(fgsea)
library(corrplot)
library(naniar)

library(EnhancedVolcano)
library(pheatmap)

library(DEqMS)
library(limma)

library(factoextra)
library(FactoMineR)
library(ggpubr)
library(RColorBrewer)
library(PCAtools)

library(ggVennDiagram)
library(reshape2)

```

# 2. Import the quantitative proteomics data

The quantitative proteomics output is imported from a tab-delimited text file.

```{r}
data <- read.delim(
  "20231221_MK_Volta_MVL_FAIMS_2CV_Easy15_140min_1785_DDA_500ng_protein.txt",
  sep = "\t",
  header = TRUE,
  check.names = FALSE
)

colnames(data)

```

The dataset contains protein identifiers, quantitative abundance measurements, and additional information generated during the proteomics workflow.

The column names are inspected to identify the quantitative measurements and verify the structure of the imported dataset.

# 3. Identify quantitative protein measurements

The quantitative abundance columns are identified programmatically based on the presence of "Normalized" in their names.

```{r}
find_quan <- function(df) {

  quan_columns <- character()

  for (cname in colnames(df)) {

    if (grepl("Normalized", cname)) {
      quan_columns <- c(quan_columns, cname)
    }

  }

  return(quan_columns)
}

quan2_columns <- find_quan(data)

quan2_columns

```

Only the protein accession and quantitative abundance measurements are retained for downstream analysis.

```{r}
dfWide <- data %>%
  subset(
    select = c("Accession", quan2_columns)
  )

```

The resulting matrix contains proteins as rows and samples as columns.

# 4. Log2 transformation

Protein abundance measurements are often highly skewed and may span several orders of magnitude. A log2 transformation is therefore applied to make the abundance distributions more suitable for statistical modelling.

```{r}

rownames(dfWide) <- dfWide$Accession

dfWide$Accession <- NULL

dfWide <- log2(dfWide)

colnames(dfWide)

```

The protein accession is retained as the row identifier throughout the analysis.

The log2 scale also provides an intuitive interpretation of fold changes. A log2 fold change of 1 corresponds to a 2-fold increase, while a log2 fold change of -1 corresponds to a 2-fold decrease.

# 5. Matching sample identifiers

The sample names in the quantitative matrix are matched to the identifiers defined in the experimental design.

```{r}
ED <- read.delim(
  "experimental_design_normalized.txt"
)

ED


name_mapping <- setNames(
  ED$Name,
  ED$SampleID
)


dfWide <- dfWide %>%
  rename_with(
    ~ name_mapping[.],
    .cols = names(dfWide)
  )

#the quantitative matrix is reordered to match the order specified in the experimental design.
order_vector <- ED$Name

dfWide <- dfWide %>%
  select(all_of(order_vector))

head(dfWide)

colnames(dfWide)

```

# 6. Assessment of missing values

Identfied proteins and missing measurements are quantified for every sample.

```{r}
bar1 <- colSums(!is.na(dfWide))

bar2 <- colSums(is.na(dfWide))

df1 <- data.frame(t(bar1))
df2 <- data.frame(t(bar2))

pnum1 <- rbind(
  df1,
  df2
)

rownames(pnum1) <- c(
  "Number of proteins",
  "Number of NAs"
)

pnum2 <- as.matrix(pnum1)

pnum2


data_long <- melt(
  pnum2
)

p1 <- ggplot(
  data_long,
  aes(
    x = Var2,
    y = value,
    fill = forcats::fct_rev(Var1)
  )
) +
  geom_bar(
    position = "stack",
    stat = "identity"
  ) +
  xlab("Sample") +
  ylab("Count") +
  labs(
    fill = "Measurement"
  ) +
  scale_fill_brewer(
    palette = 1
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(
      angle = 90,
      hjust = 1,
      size = 12
    ),
    axis.text.y = element_text(
      size = 12
    )
  )

p1

```

# 7. Missing-value filtering

In this dataset, the observed level of missingness was considered sufficiently low to remove proteins containing missing observations rather than perform imputation.

```{r}


n_before <- nrow(dfWide)

dfWide <- na.omit(dfWide)

n_after <- nrow(dfWide)

cat(
  "Proteins before filtering:", n_before, "\n"
)

cat(
  "Proteins after filtering:", n_after, "\n"
)

cat(
  "Proteins removed:", n_before - n_after, "\n"
)

```

# 8. Sample correlation analysis

Pairwise Pearson correlations are calculated between samples using their protein abundance profiles.

```{r}
cm <- dfWide

cm <- cor(
  cm,
  method = "pearson"
)

pheatmap(
  cm,
  main = "Sample correlation matrix"
)
```

# 9. Principal Component Analysis

Principal Component Analysis (PCA) is used to investigate the major sources of variation within the proteomic dataset.

Because samples are the observations and proteins are the measured variables, the abundance matrix is transposed before PCA.

```{r}
condition <- ED$Group

sample <- ED$Name

pca_data <- prcomp(
  t(dfWide)
)

pca_data_perc <- round(
  100 *
    pca_data$sdev^2 /
    sum(pca_data$sdev^2),
  1
)

df_pca_data <- data.frame(
  PC1 = pca_data$x[, 1],
  PC2 = pca_data$x[, 2],
  sample = colnames(dfWide),
  condition = condition
)


p3 <- ggplot(
  df_pca_data,
  aes(
    PC1,
    PC2,
    color = condition
  )
) +
  geom_point(
    size = 3
  ) +
  theme_bw() +
  stat_ellipse(
    geom = "polygon",
    aes(fill = condition),
    alpha = 0.05,
    show.legend = FALSE,
    level = 0.80
  ) +
  labs(
    x = paste0(
      "PC1 ",
      pca_data_perc[1],
      " %"
    ),
    y = paste0(
      "PC2 ",
      pca_data_perc[2],
      " %"
    ),
    color = "Condition"
  )

p3

```

# 10. Differential protein abundance analysis

Differential protein abundance is assessed using the limma linear modelling framework.

```{r}
dfD <- dfWide
#Define the design vector
cond=as.factor(ED$Group)

design = model.matrix(~+cond-1)
colnames(design) = gsub("cond","",colnames(design))
design
#Make contrasts
x <- c("T30m-ctr", "T1h-ctr","T2h-ctr", "T4h-ctr", "T1h-T30m", "T2h-T30m","T4h-T30m","T2h-T1h","T4h-T1h","T4h-T2h")
contrast =  makeContrasts(contrasts=x,levels=design)
contrast
fit1 <- lmFit(dfD, design)
fit2 <- contrasts.fit(fit1,contrasts = contrast)
fit3 <- eBayes(fit2, trend=TRUE, robust=TRUE)


de1<-summary(decideTests(fit3))
de1
de2<-summary(decideTests(fit3, lfc=1))
de2



#cALCULATE PAIRWISE comparisons
T1<-topTable(fit3, coef=1, n="inf")
T2<-topTable(fit3, coef=2, n="inf")
T3<-topTable(fit3, coef=3, n="inf")
T4<-topTable(fit3, coef=4, n="inf")

```

# 11. Volcano plot: 4 hours versus control

A volcano plot is generated for the 4-hour versus control comparison.

```{R}

vol <- T4

keyvals <- ifelse(
  vol$logFC < -1 &
    vol$adj.P.Val < 0.05,
  "skyblue4",
  ifelse(
    vol$logFC > 1 &
      vol$adj.P.Val < 0.05,
    "lightcoral",
    "grey"
  )
)

keyvals[is.na(keyvals)] <- "grey"

names(keyvals)[
  keyvals == "lightcoral"
] <- "up"

names(keyvals)[
  keyvals == "grey"
] <- "NR"

names(keyvals)[
  keyvals == "skyblue4"
] <- "down"



v1 <- EnhancedVolcano(
  vol,
  lab = rownames(vol),
  x = "logFC",
  y = "adj.P.Val",
  xlab = bquote(~Log[2]~ "fold change"),
  title = "4 hours vs control",
  pCutoff = 0.05,
  FCcutoff = 1,
  pointSize = 1,
  subtitle = "Differential protein abundance",
  colCustom = keyvals,
  colAlpha = 1,
  legendPosition = "right",
  legendLabSize = 15,
  legendIconSize = 5,
  drawConnectors = TRUE,
  widthConnectors = 1,
  arrowheads = FALSE,
  gridlines.major = TRUE,
  gridlines.minor = FALSE,
  border = "full",
  borderWidth = 0.7
)

v1

```

```{R}
#How many DEGs are common between different time points (lfc>1)


library(ggVennDiagram)
library(ggplot2)


T1_sign <- filter(T1, logFC< (-1) & adj.P.Val < 0.05)
T2_sign <- filter(T2, logFC< (-1) & adj.P.Val < 0.05)
T3_sign <- filter(T3, logFC<(-1)  & adj.P.Val < 0.05)
T4_sign <- filter(T4, logFC<(-1) & adj.P.Val < 0.05)



sign<-list(T30min=rownames(T1_sign), T1h=rownames(T2_sign), T2h=rownames(T3_sign), T4h=rownames(T4_sign))

venn <- ggVennDiagram(
  sign,
  label = "count",
  set_color = "black",
  label_color = "black"
) +
  scale_fill_gradient(
    low = "white",
    high = "coral3"
)

venn

```

The Venn diagram summarizes the overlap of significantly downregulated proteins across the time course.

Proteins shared between multiple time points may represent a persistent response, whereas proteins restricted to individual time points may indicate more transient or time-specific regulation.

The temporal overlap therefore provides a complementary perspective to the individual differential abundance analyses.


#  Download GO annotations for S. aureus USA300

```{R}

library(clusterProfiler)
library(dplyr)
library(readr)
library(ggplot2)



go_url <- paste0(
  "https://rest.uniprot.org/uniprotkb/stream?",
  "compressed=false&",
  "format=tsv&",
  "query=%28organism_id%3A367830%29&",
  "fields=accession%2Cgo_id"
)

go_annotations <- read.delim(
  go_url,
  sep = "\t",
  header = TRUE,
  stringsAsFactors = FALSE
)

head(go_annotations)


# 2. Clean GO annotation table


colnames(go_annotations)

go_annotations <- go_annotations %>%
  filter(
    !is.na(Gene.Ontology.IDs),
    Gene.Ontology.IDs != ""
  )

head(go_annotations)

colnames(go_annotations)

go_annotations <- go_annotations %>%
  select(
    Entry,
    `Gene Ontology IDs`
  ) %>%
  filter(
    !is.na(`Gene Ontology IDs`)
  )


# ---------------------------------------------------------
# 3. Create protein -> GO term mapping
# ---------------------------------------------------------

go_map <- go_annotations %>%
  tidyr::separate_rows(
    `Gene Ontology IDs`,
    sep = ";"
  ) %>%
  filter(
    !is.na(`Gene Ontology IDs`),
    `Gene Ontology IDs` != ""
  ) %>%
  select(
    Entry,
    GO = `Gene Ontology IDs`
  )

head(go_map)

T4_up <- T4 %>%
  filter(
    logFC > 1,
    adj.P.Val < 0.05
  )

T4_down <- T4 %>%
  filter(
    logFC < -1,
    adj.P.Val < 0.05
  )

T4_up_ids <- rownames(T4_up)
T4_down_ids <- rownames(T4_down)

background_ids <- rownames(dfWide)

background_ids <- background_ids[
  !is.na(background_ids)
]

# ---------------------------------------------------------
# 6. GO over-representation analysis
# ---------------------------------------------------------

T4_go <- enricher(
  gene = T4_up_ids,
  universe = background_ids,
  TERM2GENE = go_map %>%
    select(
      GO,
      Entry
    ),
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05
)

T4_go

# ---------------------------------------------------------
# 7. Visualize enriched GO terms
# ---------------------------------------------------------

dotplot(
  T4_go,
  showCategory = 20
) +
  ggtitle(
    "GO Biological Process enrichment - 4 h vs control"
  )

```



